TO FIND

### 2013-05-28 ms.tex
References

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{color,lineno,setspace,multirow,kpfonts}
\usepackage{graphicx}
\usepackage[top=2.4cm,left=2.4cm,top=2.4cm,bottom=2.4cm,includefoot]{geometry}
\usepackage[style=bes]{biblatex}
\bibliography{./library}

@@

Ecology needs to move toward a more predictive approach, integrating elements of
theoretical ecology \parencite{Thuiller2013}. The metacommunity perspective
\parencite{Leibold2004a} appears naturally as the appropriate conceptual
framework to fill this challenge. The metacommunity concept builds on feedbacks
between local scale processes, such as competitive interactions and local
adaptation, and regional scale processes such as dispersal, gene flow and
speciation. It is particularly relevant to limnology, where exchanges of
organisms and nutrients affect community and ecosystem properties from the local
(e.g. vertical mixing \parencite{Ryabov2011}) to the regional (e.g. connection
of lakes \parencite{Leibold2004b, Gravel2010} scales. It emphasizes the
importance of dispersal relative to pairwise interactions in the organization of
ecological communities.

@@

Neutral theory has also been proposed as a useful tool to understand and predict
some aspects of community dynamics. It links to an old philosophical debate
between realism and instrumentalism \parencite{Wennekes2011}. Because every
ecological model is a simplification of reality, scientists have to subjectively
decide the level of details they put in, leaving out some elements they consider
unimportant. The realism perspective requires that all assumptions of the theory
must be true, while the utility of the theory is more important to
instrumentalism. The utilitarian value of a theory could either be for
understanding or for prediction (another old philosophical debate, see
\textcite{Schmueli2010}). Obviously neutral theory could only be instrumental.
The question then is if such a 'general, large-scale, but vague' theory
\parencite{Wennekes2011} is a satisfying approximation.

@@

Our main objective is to use neutral theory to stress the importance of
landscape network structure on the distribution of diversity. We refer to the
landscape organization as a \emph{spatial contingency} \parencite{Peres-Neto2013}
that could potentially affect the coexistence mechanisms at play. We will
therefore move from a perspective where dispersal is either global or spatially
explicit (e.g. over a lattice), and spatial constant, to a perspective focusing
on the variance of dispersal. A second generation of neutral models (e.g.
\textcite{Economo2008, Economo2011,Desjardins2012a,Desjardins2012b}, and even
experiments \parencite{Altermat2012}, recently introduced more realistic
landscapes and found surprising contributions of spatial contingencies. We start
with a short review of the main approaches to describe spatial networks. Then we
describe three simple toy models of metacommunity dynamics, using this
opportunity to review their assumptions and main predictions. We provide as
Supplementary Material the R scripts for the toy models and all simulations
conducted for this paper. We then conduct simple simulations of these models to
reveal with simple examples the impact of spatial network structure on diversity
distribution. We conclude with a discussion on the operationally of the
framework.

@@

A network is a discrete mathematical object made of two sets: a set of nodes
(or vertices) and a set of edges connecting the nodes \cite{new10}. The
term ``graph'' is often preferred in computer science and mathematics
\cite{gro06}, with graph algorithms being an important and active area of
research \cite{sed01}. A network is a combinatorial object: it is used to study
how discrete entities are connected and how they combine together to create
complex structures. They are used to study molecules, food webs, social
networks, or even the relationship between variables in statistics
\cite{wri21,new10}. We are especially interested in spatial networks, a special
kind of network mixing the combinatorial properties of networks with a
topological space \cite{kob94}. Thus, the vertices in a spatial graph are
embedded in some other space, most often the two or three- dimensional Euclidean
space. This object brings a rich representation to spatial ecology and is
particularly suited for systems of lakes and rivers, which can easily be
represented by vertices and edges. There are two notions of distance in spatial
networks. Euclidean distance represents the geographical distance between the
nodes $(i, j)$, i.e.: $\sqrt{(x_i - x_j)^2 + (y_i - y_ j)^2}$. Geodesic
distance is the distance in the graph space, i.e.: the length of the shortest
path \cite{dij59}. For example, two lakes could be very close on a map (short
Euclidean distance) but the geodesic distance could be great if they are not
directly linked by a river.

@@

For simulations, spatial networks can easily be generated with the random
geometric graph algorithm \cite{sed01}. In this algorithm, all nodes are
assigned to a position in some two-dimensional space, most often the unit
square. Then, all pairs of nodes within some threshold Euclidean distance $r$
are connected with an edge. The resulting networks have the desirable property
of locality: if a node $A$ is connected to two vertices $B$ and $C$. then $B$
and $C$ are more likely to be connected than two random vertices. Random
geometric networks have been extensively studied \cite{app97a,app97
b,app02a,app02b,pen03} and we provide a R function to generate them. The
position of nodes is typically random, but we could also imagine alterations
where they are either more aggregated or segregated than expected by chance
alone.

We also provide the code for a second structure that we call a random geometric
tree. The algorithm first builds a random geometric graph, then select a node
from which to start the tree. It then calculaes the the shortest path tree
\cite{dij59} from this node to all other ones. Edges that are not located along
this tree are pruned. This random geometric tree does not exactly represent
dendritic landscapes but is a convenient model to simulate lake connected by
rivers to a series of smaller lakes.

Spatial graphs are increasingly popular in spatial ecology and conservation
biology, where patterns of connections can be used to study and influence the
flow of organisms \cite{min07,fal07,min08,gar08,urb09,dal10}. In the neutral
theory, networks were pioneered by Economo and Keitt \cite{eco08, eco10}. They
used networks to study how different spatial structures influenced diversity.
They were also used to study how the spatial structure influenced nonsympatric
speciation \cite{des12,des12b}.

@@

The simplest metacommunity model is a $S$ species extension of traditional
metapopulation models \parencite(Hanski1999). The standard Levins metapopulation
model \parencite(Levins1969) describes the stochastic colonizations and
extinctions of a single species over a homogenous landscape. The basic unit is
the population. The model tracts the dynamics of occupancy (the fraction of the
landscape that is occupied) with an ordinary differential equation and therefore
assumes an infinite landscape. The simulation model we run is more realistic as
it simulates a finite number $N$ of discrete patches (or nodes in network
terminology). The rules described in the previous section were used to generate
connectivity matrices along four scenarios (Fig. 1): global dispersal (connected
graph), a lattice, a random geometric graph and a random tree graph. A patch $x$
shares $d_x$ links with neigbhouring patches (its degree). At each time step
(the simulation model is discrete in time), the probability that a colonist
coming from an occupied patch $y$ arrives at patch $x$ is $cd_y^{-1}$, where $c$
is the probability a colonization event takes place if all connected patches are
occupied. The expected probability that a colonist arrives to patch $x$ from
patch $y$ is then $C_{ixy}=cp_{iy}d_y^{-1}$, where $p_{i}y$ is the probability
that patch $y$ is occupied by species $i$. The probability that an extinction
occurs in a given patch is $e$. The Levins model is for a single species, but a
basic metacommunity patch dynamics model could be run by aggregating $S$
independent metapopulation models \parencite(Hanski1997). There are no
interactions in this simple model, which means there is no limit to local
species richness and no carrying capacity. Competitive, mutualistic and
predator-prey interactions have been added to this framework (e.g.
\textcite{Tilman1994,Holt1996,Klausmeier1998,Gravel2011b}) but we will keep this
model minimal for the sake of comparison with the neutral model.

Predictions of the patch dynamics metacommunity model are quite straightforward.
First, a fundamental result of metapopulation ecology is that persistence will
occur if colonization probability is larger than extinction probability ($c>e$).
Given that all species are the same, then we should expect the regional
diversity ($\gamma$) to be $S$ if this condition is satisfied and $0$ if not.
The situation is however more complex in spatially explicit landscapes with
complex connectivity matrices \parencite{Hanski1998}. Spatially explicit
dispersal usually reduces the occupancy and thereby the likelihood of
persistence. The second prediction is that, given spatial variation in
connectivity, there will be spatial variation in occurrence probability. Given
the above formulation of a colonization event to occur, the probability that an
empty location is colonized is $I_ix=1-\prod d_x(1-C_{ixy})$. This equation
basically tells us that the colonization probability will increase
asymptotically with the degree of a patch (because of the product). It is easy
to show from metapopulation theory that the occurrence probability in a patch is
then $p_ix=I_x(I_ix+e)^{-1}$. The feedback between local and regional dynamics
arises because all $p_ix$ from the landscape are dependent from each other.
Simulations are usually conducted to solve the model for a large landscape, but
numerical solutions are theoretically possible. The aggregation across the $S$
species of the regional species pool is obtained by taking the summation of
occurrence probabilities over all species, $s_x = \sum{p_i}$. Because in this
model all species are equal, we expect the local species richness to be a linear
function of the patch degree (number of edges). Multi-species analysis of
metapopulation models also reveals interesting predictions on other aspects of
community organization at various spatial scales such as the species-area
relationship \parencite{Hanski1997}, and proved to be useful in conservation
ecology with predictions of extinctions following habitat destruction
\parencite{Nee1994,Rybicki2013}.

@@

The model is neutral because it assumes that the probabilities of local
recruitment, immigration and mortality events are all equal across species.
Demographic stochasticity is the source of variations in abundance, but larger
disturbances could be simulated as well, as long as they hit all species with
the same probability, independently of their abundance. The fundamental feature
of neutral dynamics is therefore the ecological drift, defined as population
changes emerging from neutraly stable population dynamics. It can be measured as
the variance between replicated time series of community dynamics
\parencite{Gravel2011a}. \textcite{Hubbell2001} provides a very comprehensive
analysis of the model, with specific attention to the effect of the different
parameters on drift (and consequently variance in abundance) and time to
extinction. Despite its simplicity, the neutral model is surprisingly rich in
the predictions it makes. \textcite{Bell2001} and \textcite{Hubbell2001}
analyzed the performance of neutral models to predict species abundance
distributions, the range-abundance relationship, spatial variation in abundance,
the species-area relationship, community turnover (beta-diversity) and
co-occurrence. Recent trophic neutral models were also found to predict
realistic ecological network structures \parencite{Canard2012}. Other than the
ecological equivalence assumption, one of the most criticized aspect of neutral
models is the realism of the speciation process and the required speciation
rates to sustain species richness \parencite{Ricklefs2003,Etienne2007}. Recent
neutral models with more credible speciation models
\parencite{Rosindell2009,Desjardins2012a} revealed the difficulty to maintain
diversity in neutral models over macro-evolutionary time scales. These models
nonetheless proposed interesting predictions on endemic species richness and
island biogeography \parencite{Rosindell2011,Desjardins2012b}.

@@

The network approach to the study of spatially explicit landscapes is a major
advancement in metacommunity ecology. It is a first step to make the concept
operational because it accounts for more realistic landscape structures and
dispersal kernels. It makes a significant departure to simple island-mainland or
global dispersal approaches used previously (e.g. \parencite{Tilman1994,
Hubbell2001, Mouquet2002}). But dispersal is also spatially explicit in a
lattice model and it does not make the landscape more realistic. We believe the
fundamental contribution of this approach is the consideration of spatial
heterogeneity of dispersal. In agreement with previous theoretical
\parencite{Economo2011, Desjardins2012} and experimemental studies
\parencite{Altermat2012}, the simulations show that the degree centrality has a
significant impact on $\alpha$ diversity. Central nodes might also contribute
more to maintain $\gamma$ diversity, as they are essential for species to spread
throughout the landscape. The nodes could be potentially quantified as keystone
for the metacommunity \parencite{Mouquet2013}. Interestingly, but not
surprisingly, this effect is weaker with species sorting dynamics. We could even
hypothesize it will vanish in the extreme case of niche differentiation (with
low overlap for instance) and low mass effect. In this particular case, the
neutral versus niche comparison therefore illustrates that very strong unequal
competitive interactions could overwhelm the impact of dispersal.

@@

The multivariate variance partitioning framework originally proposed by
\textcite{Borcard1992} and further developed by \textcite{Borcard2002} has been
widely used to quantify the relative importance of species sorting and dispersal
limitations in species distribution. This framework was originally proposed to
model species distribution as a function of environmental variables, taking into
account the spatial autocorrelation of species distribution
\parencite{Leduc1992, Borcard1992, Legendre1993}. This methodology has been
widely used over the last decade as a test of the neutral theory, its underlying
assumption and a quantification of dispersal limitations (e.g.
\textcite{Svenning2004, Hardy2004, Gilbert2004, Cottenie2005}. One problem of
this approach is however that it makes a weak test of neutrality
\parencite{McGill200}, based on the description of spatial community structure,
rather than hypothesis testing. The different models we reviewed in this article
could be better employed if used to generate null expectations of species
distribution based on different hypotheses and then compare them. But most of
all, parametrized spatially explicit neutral models could be more useful if used
to predict biodiversity under different global change scenarios. For instance,
neutral models could be used to predict the consequences of habitat destruction,
fragmentation or a change in the connectivity matrix \parencite{Hubbell2008}.
The spatially explicit description of the landscape is a major improvement
toward that end, providing much flexibility in the scenarios that could be
explored.

Working with more complex representations on landscapes has several advantages.
Real landscapes are not flat geometric objects, they are highly structured and
diverse. This diverse structure has long been recognized as a key component of
diversity. In the 19th century, Moritz Wagner noted that patterns of rivers
\cite{coy04} could explain how beetles diversified. Yet, to this day, the
relationship between spatial structures and biodiversity is not resolved, but
networks provide powerful tools to analyze landscapes and generate testable
predictions on the relationship between community assembly and spatial
structures. For example, the neutral theory predict less diversity in isolated
communities and constant speciation (regardless of isolation) \cite{eco08}.
Adding the effect of gene flow changes the predictions, but the neutral theory
do not predict adaptive radiation and might thus face problems for predicting
many long-term ecological processes.

Finally, our analyses emphasize the need to expand on the canonial neutral
theory. As pointed out by \textcite{Wootoon2005}, most of the unexplained
deviation of empirical community from the prediction of accurately calibrated
neutral models can be attributed to non-competitive interactions. \textcite{Canard2012}
proposed that neutral processes can explain the network structure of trophic
interactions with a good accuracy. Incorporating reasonable complexity in the
mechanisms addressed by neutral models is not a theoreticians' exercise: it will
re-enforce the usefulness of the neutral theory as an operational concept,
specifically one that can be used to derive baseline predictions about (i) the
expected local species richness, and (ii) the expected species pool
dissimilarity at the between-site and regional scales. These predictions are the
benchmark against which empirical relev\'es of species richness and community
structure ought to be compared, and coming up with realistic parameters to
calibrate these models calls for a closer cooperation and dialogue between
theoreticians and empiricists.

@@

%========================================================%
\section*{Figure legends}

>>>>

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{color,lineno,setspace,multirow,kpfonts}
\usepackage{graphicx}
\usepackage[top=2.4cm,left=2.4cm,top=2.4cm,bottom=2.4cm,includefoot]{geometry}
\usepackage[style=bes]{biblatex}
\bibliography{library}

@@

Ecology needs to move toward a more predictive approach, integrating elements of
theoretical ecology \parencite{Thuiller2013}. The metacommunity perspective
\parencite{Leibold2004a} appears naturally as the appropriate conceptual
framework to fill this challenge. The metacommunity concept builds on feedbacks
between local scale processes, such as competitive interactions and local
adaptation, and regional scale processes such as dispersal, gene flow and
speciation. It is particularly relevant to limnology, where exchanges of
organisms and nutrients affect community and ecosystem properties from the local
(e.g. vertical mixing \parencite{Ryabov2011}) to the regional (e.g. connection
of lakes \parencite{Leibold2004b, Gravel2010a} scales. It emphasizes the
importance of dispersal relative to pairwise interactions in the organization of
ecological communities.

@@

Neutral theory has also been proposed as a useful tool to understand and predict
some aspects of community dynamics. It links to an old philosophical debate
between realism and instrumentalism \parencite{Wennekes2012}. Because every
ecological model is a simplification of reality, scientists have to subjectively
decide the level of details they put in, leaving out some elements they consider
unimportant. The realism perspective requires that all assumptions of the theory
must be true, while the utility of the theory is more important to
instrumentalism. The utilitarian value of a theory could either be for
understanding or for prediction (another old philosophical debate, see
\textcite{Shmueli2010}). Obviously neutral theory could only be instrumental.
The question then is if such a 'general, large-scale, but vague' theory
\parencite{Wennekes2012} is a satisfying approximation.

@@

Our main objective is to use neutral theory to stress the importance of
landscape network structure on the distribution of diversity. We refer to the
landscape organization as a \emph{spatial contingency} \parencite{Peres-Neto2012}
that could potentially affect the coexistence mechanisms at play. We will
therefore move from a perspective where dispersal is either global or spatially
explicit (e.g. over a lattice), and spatial constant, to a perspective focusing
on the variance of dispersal. A second generation of neutral models (e.g.
\textcite{Economo2008, Economo2011,Desjardins2012a,Desjardins2012b}), and even
experiments \parencite{Carrara2012}, recently introduced more realistic
landscapes and found surprising contributions of spatial contingencies. We start
with a short review of the main approaches to describe spatial networks. Then we
describe three simple toy models of metacommunity dynamics, using this
opportunity to review their assumptions and main predictions. We provide as
Supplementary Material the R scripts for the toy models and all simulations
conducted for this paper. We then conduct simple simulations of these models to
reveal with simple examples the impact of spatial network structure on diversity
distribution. We conclude with a discussion on the operationally of the
framework.

@@

A network is a discrete mathematical object made of two sets: a set of nodes
(or vertices) and a set of edges connecting the nodes \parencite{new10}. The
term ``graph'' is often preferred in computer science and mathematics
\parencite{gro06}, with graph algorithms being an important and active area of
research \parencite{sed01}. A network is a combinatorial object: it is used to study
how discrete entities are connected and how they combine together to create
complex structures. They are used to study molecules, food webs, social
networks, or even the relationship between variables in statistics
\parencite{wri21,new10}. We are especially interested in spatial networks, a special
kind of network mixing the combinatorial properties of networks with a
topological space \parencite{kob94}. Thus, the vertices in a spatial graph are
embedded in some other space, most often the two or three- dimensional Euclidean
space. This object brings a rich representation to spatial ecology and is
particularly suited for systems of lakes and rivers, which can easily be
represented by vertices and edges. There are two notions of distance in spatial
networks. Euclidean distance represents the geographical distance between the
nodes $(i, j)$, i.e.: $\sqrt{(x_i - x_j)^2 + (y_i - y_ j)^2}$. Geodesic
distance is the distance in the graph space, i.e.: the length of the shortest
path \parencite{dij59}. For example, two lakes could be very close on a map (short
Euclidean distance) but the geodesic distance could be great if they are not
directly linked by a river.

@@

For simulations, spatial networks can easily be generated with the random
geometric graph algorithm \parencite{sed01}. In this algorithm, all nodes are
assigned to a position in some two-dimensional space, most often the unit
square. Then, all pairs of nodes within some threshold Euclidean distance $r$
are connected with an edge. The resulting networks have the desirable property
of locality: if a node $A$ is connected to two vertices $B$ and $C$. then $B$
and $C$ are more likely to be connected than two random vertices. Random
geometric networks have been extensively studied \parencite{app97a,app97b,app02a,app02b,pen03} 
and we provide a R function to generate them. The
position of nodes is typically random, but we could also imagine alterations
where they are either more aggregated or segregated than expected by chance
alone.

We also provide the code for a second structure that we call a random geometric
tree. The algorithm first builds a random geometric graph, then select a node
from which to start the tree. It then calculaes the the shortest path tree
\parencite{dij59} from this node to all other ones. Edges that are not located along
this tree are pruned. This random geometric tree does not exactly represent
dendritic landscapes but is a convenient model to simulate lake connected by
rivers to a series of smaller lakes.

Spatial graphs are increasingly popular in spatial ecology and conservation
biology, where patterns of connections can be used to study and influence the
flow of organisms \parencite{min07,fal07,min08,gar08,urb09,dal10}. In the neutral
theory, networks were pioneered by Economo and Keitt \parencite{Economo2008, eco10}. They
used networks to study how different spatial structures influenced diversity.
They were also used to study how the spatial structure influenced nonsympatric
speciation \parencite{Desjardins2012a,Desjardins2012b}.

@@

The simplest metacommunity model is a $S$ species extension of traditional
metapopulation models \parencite(Hanski1999). The standard Levins metapopulation
model \parencite(Levins1969) describes the stochastic colonizations and
extinctions of a single species over a homogenous landscape. The basic unit is
the population. The model tracts the dynamics of occupancy (the fraction of the
landscape that is occupied) with an ordinary differential equation and therefore
assumes an infinite landscape. The simulation model we run is more realistic as
it simulates a finite number $N$ of discrete patches (or nodes in network
terminology). The rules described in the previous section were used to generate
connectivity matrices along four scenarios (Fig. 1): global dispersal (connected
graph), a lattice, a random geometric graph and a random tree graph. A patch $x$
shares $d_x$ links with neigbhouring patches (its degree). At each time step
(the simulation model is discrete in time), the probability that a colonist
coming from an occupied patch $y$ arrives at patch $x$ is $cd_y^{-1}$, where $c$
is the probability a colonization event takes place if all connected patches are
occupied. The expected probability that a colonist arrives to patch $x$ from
patch $y$ is then $C_{ixy}=cp_{iy}d_y^{-1}$, where $p_{i}y$ is the probability
that patch $y$ is occupied by species $i$. The probability that an extinction
occurs in a given patch is $e$. The Levins model is for a single species, but a
basic metacommunity patch dynamics model could be run by aggregating $S$
independent metapopulation models \parencite(Hanski1997). There are no
interactions in this simple model, which means there is no limit to local
species richness and no carrying capacity. Competitive, mutualistic and
predator-prey interactions have been added to this framework (e.g.
\textcite{Tilman1994a,Holt1996,Klausmeier1998,Gravel2011b}) but we will keep this
model minimal for the sake of comparison with the neutral model.

Predictions of the patch dynamics metacommunity model are quite straightforward.
First, a fundamental result of metapopulation ecology is that persistence will
occur if colonization probability is larger than extinction probability ($c>e$).
Given that all species are the same, then we should expect the regional
diversity ($\gamma$) to be $S$ if this condition is satisfied and $0$ if not.
The situation is however more complex in spatially explicit landscapes with
complex connectivity matrices \parencite{Hanski1998}. Spatially explicit
dispersal usually reduces the occupancy and thereby the likelihood of
persistence. The second prediction is that, given spatial variation in
connectivity, there will be spatial variation in occurrence probability. Given
the above formulation of a colonization event to occur, the probability that an
empty location is colonized is $I_ix=1-\prod d_x(1-C_{ixy})$. This equation
basically tells us that the colonization probability will increase
asymptotically with the degree of a patch (because of the product). It is easy
to show from metapopulation theory that the occurrence probability in a patch is
then $p_ix=I_x(I_ix+e)^{-1}$. The feedback between local and regional dynamics
arises because all $p_ix$ from the landscape are dependent from each other.
Simulations are usually conducted to solve the model for a large landscape, but
numerical solutions are theoretically possible. The aggregation across the $S$
species of the regional species pool is obtained by taking the summation of
occurrence probabilities over all species, $s_x = \sum{p_i}$. Because in this
model all species are equal, we expect the local species richness to be a linear
function of the patch degree (number of edges). Multi-species analysis of
metapopulation models also reveals interesting predictions on other aspects of
community organization at various spatial scales such as the species-area
relationship \parencite{Hanski1997}, and proved to be useful in conservation
ecology with predictions of extinctions following habitat destruction
\parencite{Tilman1994b,Rybicki2013}.

@@

The model is neutral because it assumes that the probabilities of local
recruitment, immigration and mortality events are all equal across species.
Demographic stochasticity is the source of variations in abundance, but larger
disturbances could be simulated as well, as long as they hit all species with
the same probability, independently of their abundance. The fundamental feature
of neutral dynamics is therefore the ecological drift, defined as population
changes emerging from neutraly stable population dynamics. It can be measured as
the variance between replicated time series of community dynamics
\parencite{Gravel2011a}. \textcite{Hubbell2001} provides a very comprehensive
analysis of the model, with specific attention to the effect of the different
parameters on drift (and consequently variance in abundance) and time to
extinction. Despite its simplicity, the neutral model is surprisingly rich in
the predictions it makes. \textcite{Bell2001} and \textcite{Hubbell2001}
analyzed the performance of neutral models to predict species abundance
distributions, the range-abundance relationship, spatial variation in abundance,
the species-area relationship, community turnover (beta-diversity) and
co-occurrence. Recent trophic neutral models were also found to predict
realistic ecological network structures \parencite{Canard2012}. Other than the
ecological equivalence assumption, one of the most criticized aspect of neutral
models is the realism of the speciation process and the required speciation
rates to sustain species richness \parencite{Ricklefs2003,Etienne2007}. Recent
neutral models with more credible speciation models
\parencite{Rosindell2010,Desjardins2012a} revealed the difficulty to maintain
diversity in neutral models over macro-evolutionary time scales. These models
nonetheless proposed interesting predictions on endemic species richness and
island biogeography \parencite{Rosindell2011,Desjardins2012b}.

@@

The network approach to the study of spatially explicit landscapes is a major
advancement in metacommunity ecology. It is a first step to make the concept
operational because it accounts for more realistic landscape structures and
dispersal kernels. It makes a significant departure to simple island-mainland or
global dispersal approaches used previously (e.g. \parencite{Tilman1994a,
Hubbell2001, Mouquet2002}). But dispersal is also spatially explicit in a
lattice model and it does not make the landscape more realistic. We believe the
fundamental contribution of this approach is the consideration of spatial
heterogeneity of dispersal. In agreement with previous theoretical
\parencite{Economo2011, Desjardins2012b} and experimemental studies
\parencite{Carrara2012}, the simulations show that the degree centrality has a
significant impact on $\alpha$ diversity. Central nodes might also contribute
more to maintain $\gamma$ diversity, as they are essential for species to spread
throughout the landscape. The nodes could be potentially quantified as keystone
for the metacommunity \parencite{Mouquet2013}. Interestingly, but not
surprisingly, this effect is weaker with species sorting dynamics. We could even
hypothesize it will vanish in the extreme case of niche differentiation (with
low overlap for instance) and low mass effect. In this particular case, the
neutral versus niche comparison therefore illustrates that very strong unequal
competitive interactions could overwhelm the impact of dispersal.

@@

The multivariate variance partitioning framework originally proposed by
\textcite{Borcard1992} and further developed by \textcite{Borcard2002} has been
widely used to quantify the relative importance of species sorting and dispersal
limitations in species distribution. This framework was originally proposed to
model species distribution as a function of environmental variables, taking into
account the spatial autocorrelation of species distribution
\parencite{Leduc1992, Borcard1992, Legendre1993}. This methodology has been
widely used over the last decade as a test of the neutral theory, its underlying
assumption and a quantification of dispersal limitations (e.g.
\textcite{Svenning2004, Hardy2004, Gilbert2004, Cottenie2005}. One problem of
this approach is however that it makes a weak test of neutrality
\parencite{McGill2003}, based on the description of spatial community structure,
rather than hypothesis testing. The different models we reviewed in this article
could be better employed if used to generate null expectations of species
distribution based on different hypotheses and then compare them. But most of
all, parametrized spatially explicit neutral models could be more useful if used
to predict biodiversity under different global change scenarios. For instance,
neutral models could be used to predict the consequences of habitat destruction,
fragmentation or a change in the connectivity matrix \parencite{Hubbell2008}.
The spatially explicit description of the landscape is a major improvement
toward that end, providing much flexibility in the scenarios that could be
explored.

Working with more complex representations on landscapes has several advantages.
Real landscapes are not flat geometric objects, they are highly structured and
diverse. This diverse structure has long been recognized as a key component of
diversity. In the 19th century, Moritz Wagner noted that patterns of rivers
\cite{coy04} could explain how beetles diversified. Yet, to this day, the
relationship between spatial structures and biodiversity is not resolved, but
networks provide powerful tools to analyze landscapes and generate testable
predictions on the relationship between community assembly and spatial
structures. For example, the neutral theory predict less diversity in isolated
communities and constant speciation (regardless of isolation) \parencite{Economo2008}.
Adding the effect of gene flow changes the predictions, but the neutral theory
do not predict adaptive radiation and might thus face problems for predicting
many long-term ecological processes.

Finally, our analyses emphasize the need to expand on the canonial neutral
theory. As pointed out by \textcite{Wootton2005}, most of the unexplained
deviation of empirical community from the prediction of accurately calibrated
neutral models can be attributed to non-competitive interactions. \textcite{Canard2012}
proposed that neutral processes can explain the network structure of trophic
interactions with a good accuracy. Incorporating reasonable complexity in the
mechanisms addressed by neutral models is not a theoreticians' exercise: it will
re-enforce the usefulness of the neutral theory as an operational concept,
specifically one that can be used to derive baseline predictions about (i) the
expected local species richness, and (ii) the expected species pool
dissimilarity at the between-site and regional scales. These predictions are the
benchmark against which empirical relev\'es of species richness and community
structure ought to be compared, and coming up with realistic parameters to
calibrate these models calls for a closer cooperation and dialogue between
theoreticians and empiricists.

@@

\newpage
%========================================================%
\section*{Figure legends}
